---
title: "Estimating 5-year breast cancer risk with BCRAT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating 5-year breast cancer risk with BCRAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(BCRAT)
```

## What this package does

`BCRAT` implements the NCI Breast Cancer Risk Assessment Tool, commonly known
as the "Gail model." Given a woman's age, age at menarche, age at first live
birth, number of prior breast biopsies (and whether any showed atypical
hyperplasia), number of first-degree relatives with breast cancer, and
race/ethnicity, it returns the **5-year probability of invasive breast
cancer**.

Two interfaces are provided: the vectorised `bcrat()` function and the
ModelsCloud prediction API (`model_run()` / `get_sample_input()` /
`get_default_input()`).

## Scoring one person

```{r}
bcrat(age = 50, race = "White", age_menarche = 13, age_first_birth = 24,
      n_biopsies = 0, n_relatives = 0)
```

Additional risk factors increase the estimate:

```{r}
bcrat(age = 55, race = "White", age_menarche = 11, age_first_birth = 22,
      n_biopsies = 2, n_relatives = 1, atypical_hyperplasia = "yes")
```

## Race and ethnicity

`race` accepts `"White"`, `"Black"`, `"Hispanic"`, `"Asian"`, or
`"American Indian or Alaska Native"` (case-insensitive; numeric codes `0`-`5`
also work). Two of these categories need one extra piece of information to
match NCI's own deployed tool exactly, because the underlying science doesn't
support a single pooled model for them:

* **Hispanic** has two entirely distinct fitted models for US-born and
  foreign-born women (not one model with a nativity adjustment) — set
  `hispanic_nativity = "us_born"` (the default) or `"foreign_born"`.
* **Asian** shares one relative-risk model across six SEER subgroups but each
  has its own baseline incidence/mortality table — set `asian_subgroup` to
  `"chinese"`, `"japanese"`, `"filipino"`, `"hawaiian"`,
  `"other_pacific_islander"`, or `"other_asian"` (the default, NCI's own
  catch-all for Asian ethnicities not individually tabulated).

```{r}
# Foreign-born Hispanic vs. the US-born default, all else equal
c(us_born      = bcrat(45, "Hispanic", 12, 26, 0, 1),
  foreign_born = bcrat(45, "Hispanic", 12, 26, 0, 1, hispanic_nativity = "foreign_born"))
```

**American Indian/Alaska Native** has no independently validated model of its
own — NCI's deployed tool substitutes the White model in full, and so does
this package:

```{r}
identical(
  bcrat(50, "White", 13, 24, 1, 0),
  bcrat(50, "American Indian or Alaska Native", 13, 24, 1, 0)
)
```

## How the score is built

The model combines a race-specific relative-risk sub-model (a small logistic
regression on menarche age, first-birth age, biopsy count, atypical
hyperplasia, and family history) with race-specific baseline hazards — SEER
invasive-breast-cancer incidence and NCHS competing (non-breast-cancer)
mortality — in a single-year discrete competing-risk life table (Gail et al.
1989, equation 1). A few nuances worth knowing:

* **Nulliparity** (`age_first_birth = "nulliparous"`) is grouped with ages
  25-29, not with the oldest age-at-first-birth category.
* **Unknown** values (`"unknown"`) are accepted for `age_menarche`,
  `age_first_birth`, `n_biopsies`, `n_relatives`, and
  `atypical_hyperplasia`, and are scored at the reference/lowest-risk
  category — exactly as NCI's own tool handles missing data.
* Some covariates are **inert for specific race models**: `age_first_birth`
  never affects the Black/CARE model (its coefficient is 0, matching the
  model's exclusion of this covariate), and `age_menarche` never affects the
  US-born Hispanic model.

```{r}
# age_first_birth doesn't change the Black estimate
c(early_birth = bcrat(45, "Black", 13, 20, 1, 1, "no"),
  late_birth  = bcrat(45, "Black", 13, 35, 1, 1, "no"))
```

## Scoring a cohort

```{r}
cohort <- data.frame(
  age             = c(50, 45, 62),
  race            = c("White", "Black", "Hispanic"),
  age_menarche    = c(13, 12, 14),
  age_first_birth = c(24, "nulliparous", 26),
  n_biopsies      = c(0, 1, 0),
  n_relatives     = c(0, 1, 1)
)
cohort$risk <- bcrat(
  age = cohort$age, race = cohort$race, age_menarche = cohort$age_menarche,
  age_first_birth = cohort$age_first_birth, n_biopsies = cohort$n_biopsies,
  n_relatives = cohort$n_relatives
)
cohort[, c("age", "race", "risk")]
```

## Using the ModelsCloud API

```{r}
get_default_input()

woman <- get_default_input()
woman$age <- 60
model_run(woman)
```

```{r}
model_run(get_sample_input())
```

## Caveats

* Valid for ages 35-85, the range NCI's own deployed tool supports (the
  papers' validation cohorts were generally 35 and older).
* American Indian/Alaska Native estimates are the White model's, substituted
  for lack of an independently validated alternative — NCI's own tool
  disclaimer says the same.
* NCI's own documentation notes that Hispanic estimates carry more
  uncertainty than White/Black estimates, and that the model "needs further
  validation for Hispanic women and other subgroups."
* For research use only; not a substitute for clinical judgement.

## References

Gail MH, Brinton LA, Byar DP, et al. Projecting individualized probabilities
of developing breast cancer for white females who are being examined
annually. *J Natl Cancer Inst.* 1989;81(24):1879-1886.
doi:[10.1093/jnci/81.24.1879](https://doi.org/10.1093/jnci/81.24.1879)

Costantino JP, Gail MH, Pee D, et al. Validation studies for models
projecting the risk of invasive and total breast cancer incidence.
*J Natl Cancer Inst.* 1999;91(18):1541-1548.
doi:[10.1093/jnci/91.18.1541](https://doi.org/10.1093/jnci/91.18.1541)

Gail MH, Costantino JP, Pee D, et al. Projecting individualized absolute
invasive breast cancer risk in African American women. *J Natl Cancer Inst.*
2007;99(23):1782-1792.
doi:[10.1093/jnci/djm223](https://doi.org/10.1093/jnci/djm223)

Matsuno RK, Costantino JP, Ziegler RG, et al. Projecting individualized
absolute invasive breast cancer risk in Asian and Pacific Islander American
women. *J Natl Cancer Inst.* 2011;103(12):951-961.
doi:[10.1093/jnci/djr154](https://doi.org/10.1093/jnci/djr154)

Banegas MP, John EM, Slattery ML, et al. Projecting individualized absolute
invasive breast cancer risk in US Hispanic women. *J Natl Cancer Inst.*
2017;109(2):djw215.
doi:[10.1093/jnci/djw215](https://doi.org/10.1093/jnci/djw215)
